\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-05 honors_thesis/ppsn_paper/paper.tex
Added another instance of the paper for PPSN

\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
}

@@

\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}

@@

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed themselves.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:griewank1}
\end{figure}

\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
runs found the global optimum, 0, while some did not.  A more accurate plot for
this function is one showing what percentage of the runs have found the global
optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
30 performing the PSO algorithm gets stuck a little less than half of the time,
while 240 particles provides enough exploration to find the global optimum
every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
particles greatly improves the chance of success with Griewank, while also
finding the global optimum in far fewer rounds of evaluations than standard
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:griewank2}
\end{figure}

@@

There are some functions for which almost no exploration needs to be done;
Sphere is an example of such a function, as would be a function where the
general area of the optimal point is known, but the exact point needs to be
found.  For such functions the best use of processors is to have a small swarm
performing speculative evaluation with our second method, where all speculative
evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, a
number of particles greater than 30 but less than 240 has ``enough''
exploration and finds the optimal value every time.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome.  It is
likely that the best thing to do with those functions is to use all processors
to add particles to the swarm and do no speculative evaluation.

>>>>

\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
%\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
%of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
%(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
}

@@

%\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
\author{Author names omitted for review}

@@

\category{G.1.6}{Numerical Analysis}{Optimization}

\terms{Algorithms}

\keywords{Particle swarm optimization; speculative evaluation; parallel
optimization}

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that both SEPSO algorithms
find the optimum faster than the original PSO.  However, because the swarm size
is so small, SEPSO gets stuck a little less than half of the time.
Surpisingly, SEPSO Pick Best improves the chance of success with Griewank,
while also finding the optimum faster.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, showing percent successful by each iteration instead of average
  function value.}
  \label{fig:griewank1}
\end{figure}

Because our speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{griewank2} shows the results.
\fig{griewank2} is very similar to \fig{griewank1}, except the speculative
algorithms find the global optimum essentially 100\% of the time.  Thus with a
few more processors, our algorithms find the optimum two to three times faster
than the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

@@

There are some functions for which very little exploration needs to be done;
Sphere is an example of such a function.  For such functions the best use of
processors is to have a small swarm performing speculative evaluation with our
second method, where all speculative evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, 30
particles is not ``enough'' exploration, but 100 is.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome and how
many processors are available.  It is likely that the best thing to do with
those functions is to use all processors to add particles to the swarm and do
no speculative evaluation.
